Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-15 paper/sections/discussion.rmd
moved everything to R Mardkown and now template works out well with rticles

# Discussion {#sec-discussion}

Our results in @sec-empirical-2 indicate that state-of-the-art approaches to Algorithmic Recourse induce substantial domain and model shift, if implemented at scale in practice. These induced shifts can and should be considered as an (expected) external cost of individual recourse. While they do not affect the individual directly as long as we look at the individual in isolation, they can been seen to affect the broader group of stakeholders in automated data-driven decision-making. We have seen, for example, that out-of-sample model performance generally deteriorates in our simulation experiments. In practice, this can be seen as a cost to model owners, that is the group of stakeholders using the model as decision-making tool. As we have set out in the introduction, these model owners will generally be unwilling to carry that cost, and hence can be expected to stop offering recourse to individuals altogether. This in turn is costly to those individuals that would otherwise derive utility from being offered recourse.

So, where does this leave us? We would argue that the expected external costs of individual recourse should be shared by all stakeholders. The most straight-forward way to achieve this is to introduce a penalty for external costs in the counterfactual search objective function, as we have set out in @eq-collective. This will on average lead to more costly counterfactual outcomes. But it may help to avoid extreme scenarios, in which minimal-cost recourse is reserved to a tiny minority of individuals. We have shown various types of shift-mitigating strategies that can be used to this end. Since all of these strategies can be seen simply as specific adaption of @eq-collective, they can be applied to any of the various counterfactual generators studied here.

>>>>

# Discussion {#discussion}

Our results in Section \@ref(empirical-2) indicate that state-of-the-art approaches to Algorithmic Recourse induce substantial domain and model shift, if implemented at scale in practice. These induced shifts can and should be considered as an (expected) external cost of individual recourse. While they do not affect the individual directly as long as we look at the individual in isolation, they can been seen to affect the broader group of stakeholders in automated data-driven decision-making. We have seen, for example, that out-of-sample model performance generally deteriorates in our simulation experiments. In practice, this can be seen as a cost to model owners, that is the group of stakeholders using the model as decision-making tool. As we have set out in the introduction, these model owners will generally be unwilling to carry that cost, and hence can be expected to stop offering recourse to individuals altogether. This in turn is costly to those individuals that would otherwise derive utility from being offered recourse.

So, where does this leave us? We would argue that the expected external costs of individual recourse should be shared by all stakeholders. The most straight-forward way to achieve this is to introduce a penalty for external costs in the counterfactual search objective function, as we have set out in Equation \@ref(eq:collective). This will on average lead to more costly counterfactual outcomes. But it may help to avoid extreme scenarios, in which minimal-cost recourse is reserved to a tiny minority of individuals. We have shown various types of shift-mitigating strategies that can be used to this end. Since all of these strategies can be seen simply as specific adaption of Equation \@ref(eq:collective), they can be applied to any of the various counterfactual generators studied here.

### 2022-08-15 paper/sections/empirical.rmd
moved everything to R Mardkown and now template works out well with rticles

# Experiment Setup {#sec-empirical}

This section presents the exact ingredients and parameter choices describing the simulation experiments we ran to produce the findings presented in the next section (@sec-empirical-2). For convenience, we use Algorithm \ref{algo-experiment} as a template to guide us through this section. A few high-level details upfront: each experiment is run for a total of $T=50$ rounds, where in each round we provide recourse to five percent of all individuals in the non-target class, so $B_t=0.05 * N_t^{\mathcal{D}_0}$^[As mentioned in the previous section, we end up providing recourse to a total of $\approx50\%$ by the end of round $T=50$.]. All classifiers and generative models are retrained for 10 epochs in each round $t$ of the experiment. Rather than retraining models from scratch, we initialize all parameters at their previous levels ($t-1$) and compute backpropagate for 10 epochs using the new training data as inputs into the existing model. Evaluation metrics are computed and stored every 10 rounds. To account for noise each individual experiment is repeat five times.

## $M$ -- Classifiers and Generative Models {#sec-empirical-classifiers}

@@

::: {#tbl-panel layout-ncol=1}

| | Hidden Dim. | Hidden Layers | Batch | Dropout |
|------|------|------|-----|-----|
| Synthetic | 32 | 1 | - | - |
| Real-World | 32 | 2 | 50 | 0.25 |

: MLP {#tbl-mlp}

|  | Hidden Dim. | Epochs |
|------|------|------|
| Synthetic | 2 | 100 | 
| Real-World | 8 | 250 |

: Variational Autoencoder {#tbl-vae}

Model Architectures
:::

## $\mathcal{D}$ -- Data {#sec-empirical-data}

@@

```{julia}
using Plots, PlotThemes
theme(:wong)
catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic()
function plot_data(data,title)
    plt = plot(title=title)
    scatter!(data)
    return plt
end
plts = [plot_data(data,name) for (name, data) in catalogue]
plt = plot(plts..., layout=(1,4), size=(850,200))
savefig(plt, "dev/paper/www/synthetic_data.png")
```

We use four synthetic binary classification datasets consisting of 1000 samples each. The datasets are presented in @fig-synthetic-data (see also Appendix A for a formal description). Samples from the negative class are marked in blue while samples of the positive class are marked in orange.

![Synthetic classification datasets used in our experiments.](www/synthetic_data.png){#fig-synthetic-data fig.pos="h" width="8cm" height="2cm"}

@@

We use three different real-world datasets from the Finance and Economics domain, all of which are tabular and can be used for binary classification. Firstly, we use the **Give Me Some Credit** dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@gmsc_data]. Originally consisting of 250,000 instances with 11 numerical attributes. Secondly, we use the **UCI defaultCredit** dataset [@yeh2009comparisons], a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable, whether credit card clients default on their payment. In its raw form it consists of 23 explanatory variables - 4 categorical features relating to demographic attributes^[These have been ommitted from the analysis. See @sec-limit-data for details.] and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding. Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example @pawelczyk2021carla, @joshi2019towards and @ustun2019actionable), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.

@@

All generators introduced earlier are included in the experiments: Wachter [@wachter2017counterfactual], REVISE [@joshi2019towards], CLUE [@antoran2020getting], DiCE [@mothilal2020explaining] and Greedy [@schut2021generating]. In addition, we introduce two new generators in @sec-empirical-2-mitigate that directly address the issue of endogenous domain and model shifts. We also test to what extent it may be beneficial to combine ideas underlying the various generators.

>>>>

# Experiment Setup {#empirical}

This section presents the exact ingredients and parameter choices describing the simulation experiments we ran to produce the findings presented in the next section (\@ref(empirical-2)). For convenience, we use Algorithm \ref{algo-experiment} as a template to guide us through this section. A few high-level details upfront: each experiment is run for a total of $T=50$ rounds, where in each round we provide recourse to five percent of all individuals in the non-target class, so $B_t=0.05 * N_t^{\mathcal{D}_0}$^[As mentioned in the previous section, we end up providing recourse to a total of $\approx50\%$ by the end of round $T=50$.]. All classifiers and generative models are retrained for 10 epochs in each round $t$ of the experiment. Rather than retraining models from scratch, we initialize all parameters at their previous levels ($t-1$) and compute backpropagate for 10 epochs using the new training data as inputs into the existing model. Evaluation metrics are computed and stored every 10 rounds. To account for noise each individual experiment is repeat five times.

## $M$ -- Classifiers and Generative Models {#empirical-classifiers}

@@

<!-- | | Hidden Dim. | Hidden Layers | Batch | Dropout | -->
<!-- |------|------|------|-----|-----| -->
<!-- | Synthetic | 32 | 1 | - | - | -->
<!-- | Real-World | 32 | 2 | 50 | 0.25 | -->

<!-- : MLP {#tbl-mlp}  -->

<!-- |  | Hidden Dim. | Epochs | -->
<!-- |------|------|------| -->
<!-- | Synthetic | 2 | 100 |  -->
<!-- | Real-World | 8 | 250 |  -->

<!-- : Variational Autoencoder {#tbl-vae} -->

## $\mathcal{D}$ -- Data {#empirical-data}

@@

```{julia, eval=FALSE, echo=FALSE}
using Plots, PlotThemes
theme(:wong)
catalogue = AlgorithmicRecourseDynamics.Data.load_synthetic()
function plot_data(data,title)
    plt = plot(title=title)
    scatter!(data)
    return plt
end
plts = [plot_data(data,name) for (name, data) in catalogue]
plt = plot(plts..., layout=(1,4), size=(850,200))
savefig(plt, "dev/paper/www/synthetic_data.png")
```

We use four synthetic binary classification datasets consisting of 1000 samples each. The datasets are presented in Figure \@ref(fig:synthetic-data) (see also Appendix A for a formal description). Samples from the negative class are marked in blue while samples of the positive class are marked in orange.

```{r synthetic-data, fig.cap="Synthetic classification datasets used in our experiments."}
knitr::include_graphics("www/synthetic_data.png")
```

@@

We use three different real-world datasets from the Finance and Economics domain, all of which are tabular and can be used for binary classification. Firstly, we use the **Give Me Some Credit** dataset which was open-sourced on Kaggle for the task to predict whether a borrower is likely to experience financial difficulties in the next two years [@gmsc_data]. Originally consisting of 250,000 instances with 11 numerical attributes. Secondly, we use the **UCI defaultCredit** dataset [@yeh2009comparisons], a benchmark dataset that can be used to train binary classifiers to predict the binary outcome variable, whether credit card clients default on their payment. In its raw form it consists of 23 explanatory variables - 4 categorical features relating to demographic attributes^[These have been ommitted from the analysis. See Section \@ref(limit-data) for details.] and 19 continuous features largely relating to individuals' payment histories and amount of credit outstanding. Both of these datasets have been used in the literature on Algorithmic Recourse before (see for example @pawelczyk2021carla, @joshi2019towards and @ustun2019actionable), presumably because they constitute real-world classification tasks involving individuals that compete for access to credit.

@@

All generators introduced earlier are included in the experiments: Wachter [@wachter2017counterfactual], REVISE [@joshi2019towards], CLUE [@antoran2020getting], DiCE [@mothilal2020explaining] and Greedy [@schut2021generating]. In addition, we introduce two new generators in Section \@ref(mitigate) that directly address the issue of endogenous domain and model shifts. We also test to what extent it may be beneficial to combine ideas underlying the various generators.

### 2022-08-15 paper/sections/empirical_2.rmd
moved everything to R Mardkown and now template works out well with rticles

# Experiments {#sec-empirical-2}

@@

Granular tables and visualizations summarizing all of our experimental results are published in the supplementary appendix. Here we will zoom in on illustrative examples that nicely reflect some of the broader patterns we have observed and listed above. @fig-syn presents a small subset of our results for the synthetic dataset with overlapping classes. It shows the resulting values for some of our evaluation metrics at the end of the experiment, so after all $T=50$ rounds, along with error bars indicating the variation across folds. The top row shows the estimated domain shifts. While it is not straight-forward to interpret the exact magnitude of MMD, we can see clearly that the values are different from zero and there is essentially no variation across our five folds. With respect to the domain shifts, the Greedy generator actually induces the smallest shifts. In general, we have observed the opposite. The second row shows the estimated model shifts, where here we have used the meshgrid approach explained earlier. As with the domain shifts, the observed values are clearly different from zero and variation across folds is once again small. In this case, the results for this particular dataset very much reflect the broader patterns we have observed: Latent Space generators induce the smallest shifts, followed by DiCE, then Wachter and finally Greedy. The same broad pattern also emerges in the bottom row: we observe the smallest deterioration in model performance for Latent Space generators, albeit still a reduction in the F-Score of around 5 percentage points on average.

![Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a meshgrid of points. Error bars indicate the standard deviation across folds.](www/synthetic_results.png){#fig-syn fig.pos="h" width="9cm" height="9cm"}

@@

![PLACEHOLDER: Results for real-world as table.](www/placeholder.png){#fig-real fig.pos="h" width="9cm" height="9cm"}

>>>>

# Experiments {#empirical-2}

@@

Granular tables and visualizations summarizing all of our experimental results are published in the supplementary appendix. Here we will zoom in on illustrative examples that nicely reflect some of the broader patterns we have observed and listed above. Figure \@ref(fig:syn) presents a small subset of our results for the synthetic dataset with overlapping classes. It shows the resulting values for some of our evaluation metrics at the end of the experiment, so after all $T=50$ rounds, along with error bars indicating the variation across folds. The top row shows the estimated domain shifts. While it is not straight-forward to interpret the exact magnitude of MMD, we can see clearly that the values are different from zero and there is essentially no variation across our five folds. With respect to the domain shifts, the Greedy generator actually induces the smallest shifts. In general, we have observed the opposite. The second row shows the estimated model shifts, where here we have used the meshgrid approach explained earlier. As with the domain shifts, the observed values are clearly different from zero and variation across folds is once again small. In this case, the results for this particular dataset very much reflect the broader patterns we have observed: Latent Space generators induce the smallest shifts, followed by DiCE, then Wachter and finally Greedy. The same broad pattern also emerges in the bottom row: we observe the smallest deterioration in model performance for Latent Space generators, albeit still a reduction in the F-Score of around 5 percentage points on average.

```{r syn, fig.cap="Results for synthetic data with overlapping classes. The shown model MMD (PP MMD) was computed over a meshgrid of points. Error bars indicate the standard deviation across folds."}
knitr::include_graphics("www/synthetic_results.png")
```

@@

```{r real, fig.cap="Results for real-world datasets."}
knitr::include_graphics("www/real_world_results.png")
```

### 2022-08-15 paper/sections/introduction.rmd
moved everything to R Mardkown and now template works out well with rticles

# Introduction {#sec-intro}

@@

@fig-poc illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by @wachter2017counterfactual: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c). As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d). We refer to these types of dynamics as **endogenous** because they are induced by the implementation of recourse itself. The term **macrodynamics** is borrowed from the economics literature and used to describe processes involving whole groups or societies.

![Dynamics in Algorithmic Recourse: we have a simple linear classifier trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a noticable domain shift (b); as the classifier is retrained we observe a corresponding model shift (c); as this process is repeated, the decision boundary moves away from the target class (d).](www/poc.png){#fig-poc fig.pos="h" width=45%}

We think that these types of endogenous dynamics may be problematic and warrant our attention. Firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse. Secondly and relatedly, we observe in @fig-poc that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the consumer credit example, individuals that previously would have been denied credit based on their input features are suddenly considered as creditworthy. Average default risk across all borrowers can therefore be expected to increase. Conversely, lenders that anticipate such dynamics may choose to refrain from offering recourse (and hence credit) to more than just a tiny share of individuals. In that latter and perhaps more likely scenario, the probability of being offered recourse decreases with every individual that implements recourse: in other words, the actions of first-movers exert a negative externality on future would-be borrowers.

@@

The remainder of the paper is structured as follows: @sec-related places our work in the broader context of related literature. @sec-method posits a generalized framework for gradient-based Algorithmic Recourse and introduces the notion of hidden external costs. @sec-method-2 sets out our experimental framework for modeling endogenous macrodynamics in AR. @sec-empirical presents our experimental setup, including the data, classifiers and counterfactual generators that we have employed. @sec-empirical-2 presents the results for synthetic and real-world datasets along with evidence for the effectiveness of various proposed mitigation strategies. Our findings are then discussed in the broader context of the literature in @sec-discussion, before pointing to some of the limitations of our work as well as avenues for future research in @sec-limit. Finally, @sec-conclusion concludes.

>>>>

# Introduction {#intro}

@@

Figure \@ref(fig:poc) illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by @wachter2017counterfactual: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c). As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d). We refer to these types of dynamics as **endogenous** because they are induced by the implementation of recourse itself. The term **macrodynamics** is borrowed from the economics literature and used to describe processes involving whole groups or societies.

```{r poc, fig.cap="Dynamics in Algorithmic Recourse: we have a simple linear classifier trained for binary classification (a); the implementation of AR for a random subset of individuals leads to a noticable domain shift (b); as the classifier is retrained we observe a corresponding model shift (c); as this process is repeated, the decision boundary moves away from the target class (d)."}
knitr::include_graphics("www/poc.png")
```

We think that these types of endogenous dynamics may be problematic and warrant our attention. Firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse. Secondly and relatedly, we observe in Figure \@ref(fig:poc) that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter: in the consumer credit example, individuals that previously would have been denied credit based on their input features are suddenly considered as creditworthy. Average default risk across all borrowers can therefore be expected to increase. Conversely, lenders that anticipate such dynamics may choose to refrain from offering recourse (and hence credit) to more than just a tiny share of individuals. In that latter and perhaps more likely scenario, the probability of being offered recourse decreases with every individual that implements recourse: in other words, the actions of first-movers exert a negative externality on future would-be borrowers.

@@

The remainder of the paper is structured as follows: Section \@ref(related) places our work in the broader context of related literature. Section \@ref(method) posits a generalized framework for gradient-based Algorithmic Recourse and introduces the notion of hidden external costs. Section \@ref(method-2) sets out our experimental framework for modeling endogenous macrodynamics in AR. Section \@ref(empirical) presents our experimental setup, including the data, classifiers and counterfactual generators that we have employed. Section \@ref(empirical-2)  presents the results for synthetic and real-world datasets along with evidence for the effectiveness of various proposed mitigation strategies. Our findings are then discussed in the broader context of the literature in Section \@ref(discussion), before pointing to some of the limitations of our work as well as avenues for future research in Section \@ref(limit). Finally, Section \@ref(conclusion) concludes.

### 2022-08-15 paper/sections/methodology.rmd
moved everything to R Mardkown and now template works out well with rticles

# Gradient-Based Recourse Revisited {#sec-method}

In this section we first set out a generalized framework for gradient-based counterfactual search in @sec-method-general that encapsulates the various individual recourse methods we have chosen to use in our experiments. We then introduce the notion of a hidden external cost in algorithmic recourse and extend the existing framework to explicitly address this cost in the counterfactual search objective.

## From individual recourse ... {#sec-method-general}

We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method - Wachter; secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work. In particular, we include include the following generators in our simulation experiments below: **REVISE** [@joshi2019towards], **CLUE** [@antoran2020getting], **DiCE** [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating]. Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models. We hypothesize that generating more **meaningful** counterfactuals should mitigate the endogenous dynamics illustrated in @fig-poc in @sec-intro. This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data. All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.

As we will see next, all of them can be described by the following generalized form of @eq-solution:

$$
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)}+ \lambda {h(f(s_k^\prime)) }  \right\}
\end{aligned}
$$ {#eq-general}

Here $\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K$ is the stacked $K$-dimensional array of counterfactual states  and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space. In Wachter, the state space is the feature space: $f$ is just the identity function and the number of counterfactuals $K$ is equal to one. Both REVISE and CLUE search counterfactuals in some latent embedding $S \subset \mathcal{S}$ instead of the feature space directly. The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$. In this case $f$ in @eq-general corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space. Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019towards].

CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, $M$. To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models. The approach proposed by @schut2021generating, which we shall refer to as **Greedy**, also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates. The authors show that in this setting the cost function $h(\cdot)$ in @eq-general is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Schut et al. [@schut2021generating] also show that by maximizing the predicted probability of $x^\prime$ being assigned to target class $y^*$ we also implicitly minimize predictive entropy - as in CLUE. In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as **Latent Space** generators^[In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category. They largely differ with respect to the chosen generative model: for example, the generator proposed by @dombrowski2021diffeomorphic relies on normalizing flows.].

Finally, DiCE [@mothilal2020explaining] distinguishes itself from all other generators considered here in that it aims to generate a diverse set of $K>1$ counterfactuals. Wachter et al. [@wachter2017counterfactual] show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual). In @mothilal2020explaining diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function $h(\mathbf{s}^\prime)$ and therebt produce counterfactuals $s_1, ... , s_K$ that look as different from each other as possible. The implementation of DiCE in the our library of choice - `CounterfactualExplanations.jl` - uses that exact approach. It is worth noting that for $k=1$, DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function @eq-general.

@@

All of the different approaches introduced above tackle the problem of Algorithmic Recourse from the perspective of one single individual^[DiCE recognizes that different individuals may have different objective functions, but it does not address the interdependencies between different individuals.]. To explicitly address the issue that individual recourse may affect the outcome and prospect of other individuals, we propose to extend @eq-general as follows:

$$
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  
\end{aligned}
$$ {#eq-collective}

Here $h(f(s_k^\prime))$ denotes the proxy for private costs faced by the individual as before and $\lambda_1$ governs to what extent that private cost ought to be penalized. The newly introduced term $g(f(s_k^\prime))$ is meant to capture and address external costs incurred by the collective of individuals in response to changes in $\mathbf{s}^\prime$. The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient [@pindyck2014microeconomics]. We think that this concept describes the endogenous dynamics of algorithmic recourse oberserved here very well. As with individual recourse, the exact choice of $g(\cdot)$ is not obvious, nor do we intend to provide a definite answer in this work, if such even exists. That being said, we do propose a few potential mitigation strategies in @sec-empirical-2-mitigate.

>>>>

# Gradient-Based Recourse Revisited {#method}

In this section we first set out a generalized framework for gradient-based counterfactual search in Section \@ref(method-general) that encapsulates the various individual recourse methods we have chosen to use in our experiments. We then introduce the notion of a hidden external cost in algorithmic recourse and extend the existing framework to explicitly address this cost in the counterfactual search objective.

## From individual recourse ... {#method-general}

We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method - Wachter; secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work. In particular, we include include the following generators in our simulation experiments below: **REVISE** [@joshi2019towards], **CLUE** [@antoran2020getting], **DiCE** [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating]. Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models. We hypothesize that generating more **meaningful** counterfactuals should mitigate the endogenous dynamics illustrated in Figure \@ref(fig:poc) in Section \@ref(intro). This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data. All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.

As we will see next, all of them can be described by the following generalized form of Equation \@ref(eq:general):

\begin{equation}
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)}+ \lambda {h(f(s_k^\prime)) }  \right\} (\#eq:general)
\end{aligned} 
\end{equation}

Here $\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K$ is the stacked $K$-dimensional array of counterfactual states  and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space. In Wachter, the state space is the feature space: $f$ is just the identity function and the number of counterfactuals $K$ is equal to one. Both REVISE and CLUE search counterfactuals in some latent embedding $S \subset \mathcal{S}$ instead of the feature space directly. The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$. In this case $f$ in Equation \@ref(eq:general) corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space. Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019towards].

CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, $M$. To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models. The approach proposed by @schut2021generating, which we shall refer to as **Greedy**, also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates. The authors show that in this setting the cost function $h(\cdot)$ in Equation \@ref(eq:general) is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Schut et al. [@schut2021generating] also show that by maximizing the predicted probability of $x^\prime$ being assigned to target class $y^*$ we also implicitly minimize predictive entropy - as in CLUE. In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as **Latent Space** generators^[In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category. They largely differ with respect to the chosen generative model: for example, the generator proposed by @dombrowski2021diffeomorphic relies on normalizing flows.].

Finally, DiCE [@mothilal2020explaining] distinguishes itself from all other generators considered here in that it aims to generate a diverse set of $K>1$ counterfactuals. Wachter et al. [@wachter2017counterfactual] show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual). In @mothilal2020explaining diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function $h(\mathbf{s}^\prime)$ and therebt produce counterfactuals $s_1, ... , s_K$ that look as different from each other as possible. The implementation of DiCE in the our library of choice - `CounterfactualExplanations.jl` - uses that exact approach. It is worth noting that for $k=1$, DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \@ref(eq:general).

@@

All of the different approaches introduced above tackle the problem of Algorithmic Recourse from the perspective of one single individual^[DiCE recognizes that different individuals may have different objective functions, but it does not address the interdependencies between different individuals.]. To explicitly address the issue that individual recourse may affect the outcome and prospect of other individuals, we propose to extend Equation \@ref(eq:general) as follows:

\begin{equation}
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  (\#eq:collective)
\end{aligned} 
\end{equation}

Here $h(f(s_k^\prime))$ denotes the proxy for private costs faced by the individual as before and $\lambda_1$ governs to what extent that private cost ought to be penalized. The newly introduced term $g(f(s_k^\prime))$ is meant to capture and address external costs incurred by the collective of individuals in response to changes in $\mathbf{s}^\prime$. The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient [@pindyck2014microeconomics]. We think that this concept describes the endogenous dynamics of algorithmic recourse oberserved here very well. As with individual recourse, the exact choice of $g(\cdot)$ is not obvious, nor do we intend to provide a definite answer in this work, if such even exists. That being said, we do propose a few potential mitigation strategies in Section \@ref(mitigate).
